---
title: "Flight Cancellation Classifier"
output: 
  prettydoc::html_pretty:
    theme: Architect
    highlight: github
    toc: true
    toc_depth: 2
---

Kaggle site: https://www.kaggle.com/c/flights-status

## 1. Reading Datasets and Loading Libraries 

```{r}
library(stringi)
library(dplyr)
library(randomForest)
library(ggplot2)
library(ggthemes)
```

Import Data
```{r}
flight.training<- read.csv("~/R Projects/Flight-Cancel-Classifier/data/flight.training.csv")
# original dim: 69225    46
flight.testing<- read.csv("~/R Projects/Flight-Cancel-Classifier/data/flight.testing.csv")
airport.info<- read.csv("~/R Projects/Flight-Cancel-Classifier/data/training.airport.info.csv")
delay.date<- read.csv("~/R Projects/Flight-Cancel-Classifier/data/delay.date.csv")
cancel.date<- read.csv("~/R Projects/Flight-Cancel-Classifier/data/cancel.date.csv")
delay.airport<- read.csv("~/R Projects/Flight-Cancel-Classifier/data/delay.csv")
```

## 2. Data Cleaning and Variable Transformation 

### Original Training and Testing Datasets
```{r}
#Cleaning Destnation City Airport Names
flight.training$d.state<- stri_sub(flight.training$Destination_city, -2,-1)
flight.testing$d.state<- stri_sub(flight.testing$Destination_city, -2,-1)
biggest<- c("DEN","DFW","IAD","MCO","IAH","SLC","ORD","SFO","JFK","DTW")

# Identifying the top 10 Busiest Aiports in the U.S.
flight.training$biggest2<- ifelse(flight.training$Origin_airport %in% biggest, 1,0)
flight.testing$biggest2<- ifelse(flight.testing$Origin_airport %in% biggest, 1,0)
```

### Airport Information Dataset
```{r}
airport.info<- airport.info[,-1]
colnames(airport.info)<- c("Origin_airport", "Departure", "Arrival", "Total.Ops", "Departure.Seats", "Ave.Dep.Seats", "Arrival.Seats", "Ave.Arr.Seats")
airport.info<- airport.info[c(-1:-5, -239),]
airport.info$Origin_airport<- stri_sub(airport.info$Origin_airport, 1, 3)
for(i in 2:8){
  airport.info[,i]<- as.numeric(airport.info[,i])
}
flight.training<- left_join(flight.training, airport.info, "Origin_airport")
flight.testing<- left_join(flight.testing, airport.info, "Origin_airport")
flight.training$tot<- flight.training$Arrival.Seats+flight.training$Departure.Seats
flight.testing$tot<- flight.testing$Arrival.Seats+flight.testing$Departure.Seats
flight.training$to<- flight.training$Arrival.Seats+flight.training$Departure.Seats
flight.testing$to<- flight.testing$Arrival.Seats+flight.testing$Departure.Seats
```

### Delay and Cancellation Datasets

```{r}
flight.training$Date<- paste(flight.training$MONTH, flight.training$DAY)
flight.training$Date<- gsub("[[:space:]]","",flight.training$Date)
delay.date<- delay.date[-91:-95,]
delay.date$Date<- gsub("/","\\",delay.date$Date)
flight.training<- left_join(flight.training, delay.date, "Date")
flight.testing$Date<- paste(flight.testing$MONTH, flight.testing$DAY)
flight.testing$Date<- gsub("[[:space:]]","",flight.testing$Date)
flight.testing<- left_join(flight.testing, delay.date, "Date")

cancel.date$Date<- gsub("/","\\",cancel.date$Date)
flight.training<- left_join(flight.training, cancel.date, "Date")
flight.testing<- left_join(flight.testing, cancel.date, "Date")

delay.airport<- delay.airport[,c(1,7)]
names(delay.airport)<- c("Destination_airport", "arrival.delay")
flight.training<- left_join(flight.training, delay.airport, "Destination_airport")
flight.testing<- left_join(flight.testing, delay.airport, "Destination_airport")

#Impute values for Aircraft Movement
# found on wikipedia
flight.training$Aircraft.Movement[flight.training$Destination_airport=="BNA"] <- 230229
flight.training$Aircraft.Movement[flight.training$Destination_airport=="HNL"] <- 326832

```


## 3. Final Model Construction and Testing (Random Forest)

```{r}
set.seed(1234)
indi3<- sample(nrow(flight.training), nrow(flight.training)*0.7, replace = F)
tr2<- flight.training[indi3,]
te2<- flight.training[-indi3,]

# Best Model Accuracy: 0.99841 Time Elapsed: 175.369s - CPU: 3 GHz Intel Core i5
flight.m2<- randomForest(data=flight.training, as.factor(Cancelled)~ Destination_airport + SCHEDULED_DEPARTURE + biggest2 +DAY + O.City + Distance + DAY_OF_WEEK +MONTH + Org_airport_long+ FLIGHT_NUMBER + SCHEDULED_TIME + SCHEDULED_ARRIVAL + tot +X..Delayed+precent+cancellation+arrival.delay ,mtry=5,ntree=500,importance=TRUE)

flight.m3<- randomForest(data=tr2, as.factor(Cancelled)~ Destination_airport + SCHEDULED_DEPARTURE + biggest2 +DAY + O.City + Distance + DAY_OF_WEEK +MONTH + Org_airport_long+ FLIGHT_NUMBER + SCHEDULED_TIME + SCHEDULED_ARRIVAL + tot +X..Delayed+precent+cancellation+arrival.delay ,mtry=5,ntree=500,importance=TRUE)
system.time(randomForest(data=flight.training, as.factor(Cancelled)~ Destination_airport + SCHEDULED_DEPARTURE + biggest2 +DAY + O.City + Distance + DAY_OF_WEEK +MONTH + Org_airport_long+ FLIGHT_NUMBER + SCHEDULED_TIME + SCHEDULED_ARRIVAL + tot +X..Delayed+precent+cancellation+arrival.delay ,mtry=5,ntree=500,importance=TRUE))
# Most Efficient Model Accuracy: 0.99557 Time Elapsed: 1.754s - CPU: 3 GHz Intel Core i5
flight.efficient<- randomForest(data=tr2, as.factor(Cancelled)~ Destination_airport + SCHEDULED_DEPARTURE +DAY + O.City + Distance + DAY_OF_WEEK +MONTH + Org_airport_long+ FLIGHT_NUMBER + SCHEDULED_TIME + SCHEDULED_ARRIVAL ,mtry=4,ntree=10,importance=TRUE)
system.time(randomForest(data=tr2, as.factor(Cancelled)~ Destination_airport + SCHEDULED_DEPARTURE +DAY + O.City + Distance + DAY_OF_WEEK +MONTH + Org_airport_long+ FLIGHT_NUMBER + SCHEDULED_TIME + SCHEDULED_ARRIVAL ,mtry=4,ntree=10,importance=TRUE))

# Testing Accuracy 
p2<- predict(flight.m3, newdata = te2, type = "response")
pred2<- p2 == "YES"
cc<- te2$Cancelled
t1<-table(pred2, cc)
(t1[1,1]+t1[2,2])/sum(t1)
```

```{r}
p3<- predict(flight.m2, newdata = flight.testing, type = "response")
pred3<- p3 == "YES"
pred3<- as.data.frame(pred3)
pred3$Ob<- rownames(pred3)
pred3[,3]<- pred3[,1]
pred3[,1]<- pred3[,2]
pred3[,2]<- pred3[,3]
pred3<- pred3[,-3]
pred3[,2]<- gsub(FALSE, "NO", pred3[,2])
pred3[,2]<- gsub(TRUE, "YES", pred3[,2])
colnames(pred3)<- c("Ob", "Cancelled")
write.csv(pred3, "flight.pred.csv", row.names = FALSE)
```



## 4. Supporting Visualizations

```{r}
tt<- as.data.frame(prop.table(table(flight.training$Cancelled, flight.training$to)))
tt<- tt[seq(2, nrow(tt),2),]
tt$Var2<-as.numeric(as.character(tt$Var2))
ggplot(tt, aes(x=as.numeric(Var2), y=Freq))+
  geom_point()+
  theme_classic()
```

Variance Important plots
```{r}
library(forcats)
df = data.frame(flight.m2$importance)
df$Variable = factor(rownames(df))
ggplot(df)

df %>% mutate(Variable=fct_reorder(Variable,MeanDecreaseAccuracy))%>%
  ggplot(aes(x=Variable,y=MeanDecreaseAccuracy)) + geom_bar(stat="identity",fill="mediumorchid4")+ theme_bw() + theme( axis.title.y=element_blank(),axis.title.x=element_blank(),axis.text.y=element_text(face="bold",size=10), plot.title = element_text(hjust=.5)) + coord_flip()  +ggtitle("Mean Decrease in Accuracy")
# 718 514

df %>% mutate(Variable=fct_reorder(Variable,MeanDecreaseGini))%>%
  ggplot(aes(x=Variable,y=MeanDecreaseGini)) + geom_bar(stat="identity",fill="mediumorchid4")+ theme_bw() + theme( axis.title.y=element_blank(),axis.title.x=element_blank(),axis.text.y=element_text(face="bold",size=10), plot.title = element_text(hjust=.5)) + coord_flip()  +ggtitle("Mean Decrease Gini")
```

Origin_Pop
Aircraft movement
Passengers
```{r}
library(gridExtra)
A<- ggplot(flight.training,aes(x=Cancelled,y=Origin_population,fill=Cancelled))+geom_boxplot() +theme_bw()+ theme(axis.title.y=element_blank(),plot.title = element_text(hjust=.5,face="bold",size=15),legend.position = "none")+scale_fill_manual(values=c("mediumorchid4","skyblue4")) + ggtitle("Origin Population")

B<-ggplot(flight.training,aes(x=Cancelled,y=Passengers,fill=Cancelled))+geom_boxplot() +theme_bw()+ theme(axis.title.y=element_blank(),plot.title = element_text(hjust=.5,face="bold",size=15),legend.position = "none")+scale_fill_manual(values=c("mediumorchid4","skyblue4")) + ggtitle("Passengers")

C<-ggplot(flight.training,aes(x=Cancelled,y=Aircraft.Movement,,fill=Cancelled))+geom_boxplot() +theme_bw()+ theme(axis.title.y=element_blank(),plot.title = element_text(hjust=.5,face="bold",size=15),legend.position = "none")+scale_fill_manual(values=c("mediumorchid4","skyblue4")) + ggtitle("Aircraft Movement")


```




## 5. Other classifiers tried

###Logistic Regression
Features chosen step wise, started with a model with most features then took out insiginificant features
```{r}


vec = unname(apply(flight.training,2,function(x) sum(is.na(x))))
vec = vec == 0
vec = unname(vec)
ft_complete = flight.training[,vec]

ft_complete = ft_complete[,-c(1:7)]
ft_complete = ft_complete[,-24]
#take out d.state

tr2_2<- ft_complete[indi3,]
te2_2<- ft_complete[-indi3,]


LR.m1 <- glm(as.factor(Cancelled)~. -Date -SCHEDULED_ARRIVAL -tot -to -Total.Ops -Passengers -DAY_OF_WEEK -SCHEDULED_DEPARTURE -FLIGHT_NUMBER -Dest_airport_lat -Org_airport_lat -Ave.Dep.Seats -Ave.Arr.Seats -Seats -Average.1 -Actual -X..Delayed.1 -Average.Passengers -Rank.Status  -Rank -DIVERTED -Flights -Org_airport_long -Scheduled -biggest2, data=tr2_2, family="binomial")

probs = predict(LR.m1,te2_2,type="response")

preds = ifelse(probs > .5, "YES","NO")

mean(preds == te2_2$Cancelled)

summary(LR.m1)

```

Best GLM model was about 75% accurate


LDA
```{r}
library(MASS)

lda.fit = lda(factor(Cancelled)~.,data=tr2_2[,c(1:26,28:48)])
preds = predict(lda.fit,te2_2)
class = preds$class
mean(class == te2_2$Cancelled)
```

LDA got around 80%


QDA
```{r}
qda.fit = qda(factor(Cancelled)~Passengers+Distance+Org_airport_long+Actual+arrival.delay+X..On.Time.1+X..On.Time+cancellation+Dest_airport_lat+Average.1+X..Delayed.1+Delayed.1+Arrivals+Scheduled+Average+SCHEDULED_DEPARTURE+X..Delayed+Delayed+Departure+Arrival+biggest2+Rank+Average.Passengers+SCHEDULED_TIME+SCHEDULED_ARRIVAL+Dest_airport_long+FLIGHT_NUMBER+MONTH+DAY+Org_airport_lat+DAY_OF_WEEK+X..Delayed+arrival.delay+Flights+Origin_population+Seats + Destination_population,data=tr2_2)
preds = predict(qda.fit,te2_2)
class = preds$class
mean(class == te2_2$Cancelled)
```

QDA got around 80%



### KNN
Used same variables as used in logistic regression
```{r}
library(class)

classes = sapply(ft_complete,class)
classes = unname(classes)

index = classes %in% c("integer","numeric")
temp.X = ft_complete[,index]

temp.X = scale(temp.X)
temp.train.X = temp.X[indi3,]
temp.test.X = temp.X[-indi3,]

train.Y = ft_complete$Cancelled[indi3]
test.Y = ft_complete$Cancelled[-indi3]

ind = seq(1,10,1)
for(i in 1:10){
        knn.pred = knn(temp.train.X,temp.test.X,train.Y,k=ind[i])
        print(paste0("k = ",ind[i]))
        print( mean(knn.pred == test.Y) )
}
knn.pred = knn(temp.train.X,temp.test.X,train.Y,k=1)

mean(knn.pred == test.Y)
```
Best with KNN: k=1, 0.9748652
Tells us that the true decision boundary is not linear

### Boost

```{r}

set.seed(2)
boost.2 = gbm(Cancelled2~ factor(Destination_airport)  + SCHEDULED_DEPARTURE + biggest2 +DAY + factor(O.City) + Distance + DAY_OF_WEEK +MONTH + Org_airport_long+ FLIGHT_NUMBER + SCHEDULED_TIME + SCHEDULED_ARRIVAL + tot  ,data=forBoost.train,distribution="bernoulli",n.trees=10000,interaction.depth =4,shrinkage = .525)

treez = seq(8000,10000,30)

err = rep(NA,67)
for(j in 1:67){
                probsBoost = predict(boost.2, newdata=forBoost.test, n.trees=treez[j], type="response")
                sequence = seq(0,1,.001)
                acc = rep(NA,1001)
                for(i in 1:1001){
                                 predsBoost  = ifelse(probsBoost > sequence[i],1,0) 
                                 acc[i]= mean(predsBoost == forBoost.test$Cancelled2)
                }

                predsBoost  = ifelse(probsBoost > sequence[which.max(acc)],1,0) 
                err[j] = mean(predsBoost == forBoost.test$Cancelled2)
}

set.seed(2)
which.max(err)
treez[27]
probsBoost = predict(boost.2, newdata=forBoost.test, n.trees=8780, type="response")

sequence = seq(0,1,.001)
acc = rep(NA,1001)

for(i in 1:1001){
                predsBoost = ifelse(probsBoost > sequence[i],1,0)
                acc[i] = mean(predsBoost == forBoost.test$Cancelled2)
}

which.max(acc)
sequence[913]
predsBoost = ifelse(probsBoost > .912,1,0)
mean(predsBoost == forBoost.test$Cancelled2)
#0.9929218
```


